Last glacial cycle glacier erosion potential in the Alps

### 2021-05-23 articles/erosion/alps-erosion.tex
Add a paragraph to discuss erosion and sliding.

    That said, the magnitude of the modelled erosion rates can hardly be used
    as a criteria to rule out this or that erosion law for the glacial Alps.
    While sliding is here modelled by a pseudo-plastic law based on
    till dilatation under high water pressure \citep{Tulaczyk.etal.2000},
    empirical erosion rules are thought to be a proxy for glacier abrasion
    (cf.~\citealp{Hallet.1979}; and possibly plucking) of a glacier flowing
    over a hard bed. Current erosion rates in the Alps strongly vary depending
    on the substratum \citep{Steinemann.etal.2020, Steinemann.etal.2021}, and
    it is arguable that a layer of till and proglacial sediments may have
    armored the glacier bed from erosion for at least parts of the glacial
    cycle. Thus, given the lack of data to constrain the erosion rule, it can
    not be excluded that a linear or sub-linear erosion law
    \citep[e.g.,][]{Cook.etal.2020} would lead to a less localized pattern of
    erosion (Fig.~\ref{fig:powerlaws}c and~d), and a stronger correlation
    between modelled ice volume and erosion (Fig.~\ref{fig:powerlaws}k and~l).
    \todo{discuss Koppes without outliers}
    Even then, the intrinsically large variations in modelled glacier sliding
    velocities yield cumulative glacial-cycle erosion two to three orders of
    magnitude higher in the Alpine valleys than near the mountain tops
    (Fig.~\ref{fig:powerlaws}d).
    \todo{discuss sliding.}
    This result is compatible with the
    order-of-magnitude variation of glacier erosion rates observed in the
    cosmogenic nuclide record \citep{Jansen.etal.2019, Steinemann.etal.2020,
    Steinemann.etal.2021}.

>>>>

    That said, the magnitude of the modelled erosion rates can hardly be used
    as a criteria to rule out this or that erosion law for the glacial Alps.
    While sliding is here modelled by a pseudo-plastic law based on
    till dilatation under high water pressure \citep{Tulaczyk.etal.2000},
    empirical erosion rules are thought to be a proxy for glacier abrasion
    (cf.~\citealp{Hallet.1979}; and possibly plucking) of a glacier flowing
    over a hard bed. Current erosion rates in the Alps strongly vary depending
    on the substratum \citep{Steinemann.etal.2020, Steinemann.etal.2021}, and
    it is arguable that a layer of till and proglacial sediments may have
    armored the glacier bed from erosion for at least parts of the glacial
    cycle. Thus, given the lack of data to constrain the erosion rule, it can
    not be excluded that a linear or sub-linear erosion law
    \citep[e.g.,][]{Cook.etal.2020} would lead to a less localized pattern of
    erosion (Fig.~\ref{fig:powerlaws}c and~d), and a stronger correlation
    between modelled ice volume and erosion (Fig.~\ref{fig:powerlaws}k and~l).
    \todo{discuss Koppes without outliers}

    The modelled pattern of erosion potential depends on PISM glacier physics
    and sliding model parameters. In particular, the pseudo-plastic sliding law
    exponent \citep[$q=0.25$ in][]{Seguinot.etal.2018} controls the sharpness
    of the transition between adherent and decoupled basal conditions. Recent
    ensemble validation of Antarctic Ice Sheet glacial-cycle simulations
    against geological and present-day observations \citep{Albrecht.etal.2020a,
    Albrecht.etal.2020b} support a higher value of $q=0.75$, and thus a
    sliding law closer to linear, which would perhaps result in a smoother
    distribution of sliding velocity and erosion potential. Lateral stress
    gradients missing from the shallow-shelf approximation stress balance could
    also contribute to moderate sliding velocity in narrow troughs
    \citep{Herman.etal.2011, Egholm.etal.2012a, Egholm.etal.2012b,
    Pedersen.etal.2014}. \todo{check these references}.

    Because of the intrinsically large variations in modelled glacier sliding
    velocities yield cumulative glacial-cycle erosion two to three orders of
    magnitude higher in the Alpine valleys than near the mountain tops
    (Fig.~\ref{fig:powerlaws}d), we argue that the modelled highly localized
    pattern of erosion potential is qualitatively robust. This pattern, and
    the lithologic variations in bedrock erodability, may explain
    order-of-magnitude variation of glacier erosion rates observed in the
    cosmogenic nuclide record \citep{Jansen.etal.2019, Steinemann.etal.2020,
    Steinemann.etal.2021}.
